### 2020-11-15 latex/acrobot.tex
Another edit of the acrobot proof

Let us now compute the constrained dynamics under
~\eqref{eqn:acrobot-constraint}.
Note that \(dh_q = \begin{bmatrix}0 & 1\end{bmatrix}\), while
\[
    dh_{p_u} = \frac{-\bar{q}_a I}{1 + I^2 p_u^2}
    .
\]
Inserting these to ~\eqref{eqn:g-qupu}, we get the solution for \(p_a\) on the
constraint manifold:
\[
    p_a(q_u,p_u) = \frac{
        (1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
    }{ml^2(3+2c_a)(1+I^2 p_u^2)}
    .
\]
The dynamics for \(p_u\) do not contain \(p_a\), so they remain unchanged.
The constrained dynamics for \(q_u\) are given by 
\begin{equation*}
    \dot{q}_u = e_1\tpose \Minv(q) \begin{bmatrix}
                    p_u \\ p_a(q_u,p_u)
                \end{bmatrix} %\\
    ,
\end{equation*}
%        &= \frac{1}{ml^2(2-c_a^2)} 
%            \left( p_u - 
%                (1+c_a)\frac{(1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
%                }{ml^2(3+2c_a)(1+I^2 p_u^2)}
%            \right) 
%        ,
%\end{align*}
which can be simplified to 
\begin{equation*}
    \dot{q_u} = \frac{(1+I^2 p_u^2)p_u + m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }{ml^2(1+I^2 p_u^2)(3+2c_a)}
    .
\end{equation*}
Hence, the constrained dynamics for the acrobot under
~\eqref{eqn:acrobot-constraint} are
\begin{align}\label{eqn:acrobot-constrained-dynamics}
    \dot{q}_u &= \frac{(1+I^2 p_u^2)p_u + m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }
            {ml^2(1+I^2 p_u^2)(3+2c_a)}
        , \\
    \dot{p}_u &= - m g l (2s_u + s_{ua})
    . \nonumber
\end{align}

@@

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma = \mathbb{S}^1 \times \R\).
    Let \(E\) be the nominal energy when \(I = 0\):
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
        .
    \]
\begin{enumerate}
    \item There always exists \(I \in \R_{>0}\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I \in \R_{<0}\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \mathbb{S}^1 \times \R_{> 0} \rightarrow \R\) by
    \[
        b(\alpha,\mu_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\alpha^2 + 30c_\alpha(1 - c_\alpha)\right)
            - c_\alpha\mu_0^2
        \right)}{
        |\mu_0|\sqrt{\mu_0^2 - 30m^2gl^3(1 - c_\alpha)}
        }
        ,
    \]
    and fix \(\bar{\mu} > 60m^2gl^3\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\mu_0)d\sigma > 0\) for all 
    \(\mu_0 \in ]60m^2gl^3, \bar{\mu}[\).
    Then there exists \(I \in \R_{> 0}\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(0,\bar{\mu})\right\}
        .
    \]
    If instead \(I \in \R_{<0}\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}
%\footnotetext{In fact, we can be more precise than this. 
%    Letting \(\epsilon > 0\), we need 
%    \(I \leq \min\left\{I_1(\epsilon),I_2\right\}\) where
%    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
%    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
%    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
%    \right)^{-1/2}\)
%    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
%    See Apendix \ref{apx:I-bounds} for details.
%}
The region \(\mathcal{O}\) is the oscillation domain of the simple pendulum
obtained by setting \(q_a = 0\);
\(\Omega\) includes \(\mathcal{O}\), alongside all rotations of this 
pendulum with momentum less than \(\bar{\mu}\). 
Hence, the first result of Theorem \ref{thm:acrobot-energy-stabilization} states
that the acrobot will gain enough energy to begin performing giant motion;
the second result guarantees can achieve giants with momentum
\(|p_u| < \bar{\mu}\).

@@

@@

Supposing we initialize the acrobot at \((\beta,\rho) = (0,\rho_0)\),
then one full rotation amounts to \(\beta\) traversing \(2\pi\)rad in a clockwise
direction (\ie \(\beta\) going from \(0\) to \(\sign{\rho_0}2\pi\)).
We define a Poincar\'{e} section as the point reached on the \(p_u\) axis after
every rotation:
\[
    P(\rho_0) = \rho_0 + I \hat{\rho}_1(\sign{\rho_0}2\pi, \rho_0)
      + R(2\pi,I)
    .
\]
If \(|P(\rho_0)| > |\rho_0|\), then the rotation has
driven the orbit further from the origin.
In other words, \(|p_u|\) has increased across the rotation.
If this is true for any \(\rho_0\), the acrobot is gaining momentum, which in
turn means it is gaining energy.

The function \(b(\beta,\rho_0)\) is even in both its inputs, and is 
\(2\pi\)-periodic in \(\beta\).
Hence, we have 
\begin{align*}
    \hat{\rho}_1(\sign{\rho_0}2\pi, \rho_0) &=
    \int \limits_0^{\sign{\rho_0}2\pi} b(\sigma,\rho_0)d\sigma
    , \\
     &=\sign{\rho_0}\int\limits_0^{2\pi}b(\sigma,\rho_0)d\sigma
     , \\
     &= \sign{\rho_0} \hat{\rho}_1(2\pi,\rho_0)
     .
\end{align*}
Furthermore, \(\hat{\rho}_1(2\pi,\rho_0) = \hat{\rho}_1(2\pi,|\rho_0|)\).
This implies that
\[
    |P(\rho_0)| = \left|\sign{\rho_0}\left(
        |\rho_0| + I \hat{\rho}_1(2\pi,|\rho_0|)\right)
    \right|
    = P(|\rho_0|)
    .
\]
Hence, if \(P(|\rho_0|) > |\rho_0|\), then a single rotation starting at \(\rho_0\)
will drive the acrobot's orbit further from the origin.
Equivalently, one need simply guarantee that \(\hat{\rho}_1(2\pi,\rho_0) > 0\)
for all \(\rho_0 > \sqrt{60m^2gl^3}\).

Unfortunately, we cannot adimensionalize
\(b(\beta,\rho_0)\) like we did for oscillations. 
Hence, we let \(U = ]\sqrt{60m^2gl^3},\bar{\rho}[\) and assume that 
\(\hat{\rho}_1(2\pi,\rho_0) > 0 \) for all \(\rho_0 \in U\).
We conclude that, for small enough \(I\), the Poincar\'{e} section 
\(P(\rho_0)\) is monotonically diverging from 0 on \(U\)
and the acrobot is gaining momentum.

\subsection*{Energy Gain}
We will prove the acrobot gains energy on the set 
\[
    E_{\bar{\mu}} := 
    \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) < E(0,\bar{\rho}) \right\}
    .
\]
% Can we use the jacobian? The spectrum is hard to find.

Recall that on \(\mathcal{O}\) we have \(\dot{\alpha} > 0\). 
Hence, any orbit which begins and remains in \(\mathcal{O}\)
must spiral clockwise around the origin until it intersect the \(q_u\) axis.
Once it does, the Poincar\'{e} section \(P_\mathcal{O}(\cdot)\) 
demonstrates the orbit must spiral away from the origin; 
this implies there cannot be any closed orbits on \(\mathcal{O}\), and that the
origin is a repeller.

Once the orbit reaches an intersection point \(\mu_0\) on the \(q_u\)-axis such
that \(P_\mathcal{O}(\mu_0) > \pi\), the orbit will no longer return to the
\(q_u\)-axis. 
Instead, it will escape \(\mathcal{O}\) entirely.
This must always happen in a finite number of iterations of the 
Poincar\'{e} section, which means the orbit escapes \(\mathcal{O}\) in finite
time., which implies the orbit must also escape
any compact set in \(\mathcal{O}\) in finite time. 
By Definition \ref{defn:energy-gain}, the VNHC is injecting energy into the
acrobot on \(\mathcal{O}\).

If we have \(I < 0\), it is equivalent to using the constraint 
\(q_a = -\bar{q}_a \arctan(I p_u)\). 
By the same analysis, this injects energy on \(\mathcal{O}\) in negative-time, so the
constraint dissipates energy on \(\mathcal{O}\).

% TODO: W and D's boundary is two elements: the point (pi,0) and the curve of the
% homoclinic orbit connecting (pi,0) to (-pi,0). We need to show that almost
% every initial condition starting in \(mathcal{O}\) will NOT converge to (pi,0), and
% therefore exit D and enter W.

Define the set of points in \(\Gamma\) which have pseudo-radius in \(U\) by
\[
    W := \left\{(q_u,p_u) \in \Gamma \mid |\rho(q_u,p_u)| \in U \right\}
    ,
\]
as in Figure \ref{fig:acrobot-rot-W}.
Monotonicity of \(P(\rho_0)\) means there are no closed orbits of 
\((q_u,p_u)\) on \(W\). 
Furthermore, \(|p_u|\) is increasing every time the orbit hits the
\(p_u\)-axis.

>>>>

Let us now compute the constrained dynamics under
~\eqref{eqn:acrobot-constraint}.
Note that \(dh_q = \begin{bmatrix}0 & 1\end{bmatrix}\), while
\[
    dh_{p_u} = \frac{-\bar{q}_a I}{1 + I^2 p_u^2}
    .
\]
Inserting these to ~\eqref{eqn:g-qupu}, we get the solution for \(p_a\) on the
constraint manifold:
\[
    p_a(q_u,p_u) = \frac{
        (1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
    }{ml^2(3+2c_a)(1+I^2 p_u^2)}
    .
\]
The dynamics for \(p_u\) do not contain \(p_a\), so they remain unchanged.
The constrained dynamics for \(q_u\) are given by 
\begin{equation*}
    \dot{q}_u = e_1\tpose \Minv(q) \begin{bmatrix}
                    p_u \\ p_a(q_u,p_u)
                \end{bmatrix} %\\
    ,
\end{equation*}
which can be simplified into 
\begin{equation*}
    \dot{q_u} = \frac{(1+I^2 p_u^2)p_u + m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }{ml^2(1+I^2 p_u^2)(3+2c_a)}
    .
\end{equation*}
Hence, the constrained dynamics for the acrobot under
~\eqref{eqn:acrobot-constraint} are
\begin{align}\label{eqn:acrobot-constrained-dynamics}
    \dot{q}_u &= \frac{(1+I^2 p_u^2)p_u + m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }
            {ml^2(1+I^2 p_u^2)(3+2c_a)}
        , \\
    \dot{p}_u &= - m g l (2s_u + s_{ua})
    . \nonumber
\end{align}

@@

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma = \mathbb{S}^1 \times \R\).
    Let 
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
        ,
    \]
    be the energy of the simple pendulum obtained by setting \(I = 0\).
\begin{enumerate}
    \item There always exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \mathbb{S}^1 \times \R_{> 0} \rightarrow \R\) by
    \[
        b(\alpha,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\alpha^2 + 30c_\alpha(1 - c_\alpha)\right)
            - c_\alpha\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\alpha)}
        }
        ,
    \]
    and fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\mu_0)d\sigma > 0\) for all 
    \(\rho_0 \in ]\sqrt{60m^2gl^3}, \bar{\rho}[\).
    Then there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(0,\bar{\rho})\right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}
%\footnotetext{In fact, we can be more precise than this. 
%    Letting \(\epsilon > 0\), we need 
%    \(I \leq \min\left\{I_1(\epsilon),I_2\right\}\) where
%    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
%    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
%    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
%    \right)^{-1/2}\)
%    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
%    See Apendix \ref{apx:I-bounds} for details.
%}
The region \(\mathcal{O}\) is the oscillation domain of the simple pendulum
obtained by setting \(q_a = 0\);
\(\Omega\) includes \(\mathcal{O}\), alongside all rotations of this 
pendulum with momentum less than \(\bar{\mu}\). 
Hence, the first result of Theorem \ref{thm:acrobot-energy-stabilization} states
that the acrobot will gain enough energy to begin performing giant motion;
the second result guarantees can achieve giants with momentum
\(|p_u| < \bar{\mu}\).

@@

Suppose we initialize the acrobot at \((q_0,p_0) \in \mathcal{O}\).
Since on \(\mathcal{O}\) we have \(\dot{\alpha} > 0\), the orbit of the acrobot
in \(\Gamma\) spirals clockwise around the origin; 
if the spiral does not exit \(\mathcal{O}\), it must intersect the \(q_u\) axis.
Once it does, the Poincar\'{e} section \(P_\mathcal{O}(\cdot)\) tells us 
the orbit must spiral away from the origin. 
This implies there cannot be any closed orbits in \(\mathcal{O}\) and that the
origin is a repeller.

Once the orbit reaches an intersection point \((\mu_0,0)\) on the \(q_u\)-axis such
that \(P_\mathcal{O}(\mu_0) > \pi\), the orbit will no longer return to the
\(q_u\)-axis. 
Instead, it will escape \(\mathcal{O}\) entirely, which means it must 
escape any compact subset of \(\mathcal{O}\) in
finite time. 
We conclude that the acrobot is gaining energy on \(\mathcal{O}\).

Setting \(I < 0\), the Poincar\'{e} section \(P_\mathcal{O}(\mu_0)\) is now
decreasing.
Hence, the orbit in \(\mathcal{O}\) spirals towards the origin, which means the
constraint is dissipating energy on \(\mathcal{O}\).

@@

Supposing we initialize the acrobot at \((\beta,\rho) = (0,\rho_0)\),
then one full rotation amounts to \(\beta\) traversing \(2\pi\)rad in a clockwise
direction (\ie \(\beta\) going from \(0\) to \(\sign{\rho_0}2\pi\)).
We define a Poincar\'{e} section as the point reached on the \(p_u\) axis after
every rotation:
\[
    P_\mathcal{R}(\rho_0) = \rho_0 + I \hat{\rho}_1(\sign{\rho_0}2\pi, \rho_0)
      + R(2\pi,I)
    .
\]
If \(|P_\mathcal{R}(\rho_0)| > |\rho_0|\), then the rotation has
driven the orbit further from the origin.
In other words, \(|p_u|\) has increased across the rotation.
If this is true for any \(\rho_0\), the acrobot is gaining momentum, which in
turn means it is gaining energy.

The function \(b(\beta,\rho_0)\) is even in both its inputs, and is 
\(2\pi\)-periodic in \(\beta\).
Hence, we have 
\begin{align*}
    \hat{\rho}_1(\sign{\rho_0}2\pi, \rho_0) &=
    \int \limits_0^{\sign{\rho_0}2\pi} b(\sigma,\rho_0)d\sigma
    , \\
     &=\sign{\rho_0}\int\limits_0^{2\pi}b(\sigma,\rho_0)d\sigma
     , \\
     &= \sign{\rho_0} \hat{\rho}_1(2\pi,\rho_0)
     .
\end{align*}
Furthermore, \(\hat{\rho}_1(2\pi,\rho_0) = \hat{\rho}_1(2\pi,|\rho_0|)\).
This implies that
\[
    |P_\mathcal{R}(\rho_0)| = \left|\sign{\rho_0}\left(
        |\rho_0| + I \hat{\rho}_1(2\pi,|\rho_0|)\right)
    \right|
    = P_\mathcal{R}(|\rho_0|)
    .
\]
Hence, if \(P_\mathcal{R}(|\rho_0|) > |\rho_0|\) (or equivalently if
\(\hat{\rho}_1(2\pi,|\rho_0|) > 0\))
the phase \((q_u(t),p_u(t))\) initialized at \((0,\rho_0)\) 
will intersect the \(p_u\) axis at a point \(p > \rho_0\).
If we could guarantee that \(\hat{\rho}_1(2\pi,\rho_0) > 0\)
for all \(\rho_0 > \sqrt{60m^2gl^3}\), then the acrobot would be spiraling away
from the origin on \(\mathcal{R}\).

Unfortunately, we cannot adimensionalize
\(b(\beta,\rho_0)\) like we did for oscillations. 
Using the assumption , we have that 
\(\hat{\rho}_1(2\pi,\rho_0) > 0 \) for all 
\(\rho_0 \in ]\sqrt{60m^2gl^3},\rho_0[\).
We conclude that, for small enough \(I\), the Poincar\'{e} section 
\(P_\mathcal{R}(\rho_0)\) is monotonically increasing when 
\(0 < \rho_0 < \bar{\rho}\) and
monotonically decreasing when \(-\bar{\rho} < \rho_0 < 0\).

\subsection*{Proving Energy Gain}
We will prove the acrobot gains energy on the set 
\[
    \Omega := \left\{ (q_u,p_u) \in \Gamma 
     \mid E(q_u,p_u) < E(0,\bar{\rho}) \right\}
    .
\]
% Can we use the jacobian? The spectrum is hard to find.
% TODO: W and D's boundary is two elements: the point (pi,0) and the curve of the
% homoclinic orbit connecting (pi,0) to (-pi,0). We need to show that almost
% every initial condition starting in \(mathcal{O}\) will NOT converge to (pi,0), and
% therefore exit D and enter W.

Define the set of points in \(\mathcal{R}\) which have pseudo-radius in \(U\) by
\[
    W := \left\{(q_u,p_u) \in \Gamma \mid |\rho(q_u,p_u)| \in U \right\}
    ,
\]
as in Figure \ref{fig:acrobot-rot-W}.
Monotonicity of \(P(\rho_0)\) means there are no closed orbits of 
\((q_u,p_u)\) on \(W\). 
Furthermore, \(|p_u|\) is increasing every time the orbit hits the
\(p_u\)-axis.
